Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-19 manuscript/manuscript.Rmd
fight with dockerhub

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  sysdeps: 
    - apt-utils
    - grass
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

@@

@@

In addition to data from the NES and  LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related e to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing  aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo 2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we
expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal
to 10 ha (0.1 km2). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part
Fof the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input (USGS e 2018). Average link length was calculated as the total stream length in a given watershed divided v
 by the number of stream reaches after dissolving (removing) any network points that do not e occur at a stream junction. Stream density was
calculated as the length of all streams in the w
 watershed (minus artificial lines through lakes)
expressed in units of meters per hectare. Upstream lake O area was calculated as the sum of the
n
lake area in the upstream watershed expressed in square meters. Stream order ratio was defined
ly
as the number of headwater (first-order) streams in the upstream watershed of the focal lake
divided by the total number of higher order (> 1) streams (Barbera and Rosso 1989). Closest
distance to an upstream lake was defined as the shortest path-distance (rather than the straight-
line distance) to a lake upstream from the focal lake.
We calculated all connectivity metrics and lake watershed extents using the streamnet
and nhdR R packages respectively (Stachelek 2018a, Stachelek 2018b). The algorithms in the
streamnet package use the sf R package (Pebesma 2018) as well as the v.net and v.stream.order modules (Jasiewicz and Metz 2011) included in GRASS GIS (GRASS Development Team
2017). All processed connectivity data and code are available at [Zenodo DOI].

@@

where greater differences in k between R the two groups indicate greater support for a connectivity effect on P retention. Prior to model fitting, we examined the bivariate relationships between each connectivity metric, water residence i time, P loading, and other factors related to P retention
eusing Pearson’s correlation coefficients. The purpose of this exercise was twofold, to determine
the potential for collinearity among any of the variables in Eq 2 and to identify any relationships
Obetween P retention, water residence time, and other watershed and lake characteristics that were
n
not included in our model. As only one connectivity metric was l used to define g for each model
y
we did not use the results of this exercise to exclude variables from further investigation. We
quantified the relative support for an effect of each connectivity metric on P retention in more
detail by calculating the difference in the median value of the P decay parameter k between
groups (i.e. $\delta$k). We used these median k values along with median estimates of x to determine
how differences in k translated to differences in P retention (Eq 2). We judged significance by
whether or not differences in group-wise P retention were greater than the measurement
precision of P retention (> 0.01).

We fit all models in a Bayesian framework using the non-linear extension to the brms
package to interface with the Stan statistical program (Burkner 2017, Stan Development Team
2017). In both models, we set a semi-informative prior on k and x of N(1.3, 0.1) and N(0.45, 0.1)
respectively. These priors were based on the confidence intervals presented in Brett and
Benjamin (2007) and qualitatively matched those used by Cheng et al. (2010). We used the
default settings of brms and rstan to generate posterior estimates using four chains of 4,000
iterations each with no thinning and initial parameter values drawn from a uniform distribution
bounded between -2 and 2. We also used the brms package for model evaluation by computing a
FBayesian R2 following o
 the method of Gelman et al. (2017).

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each
iconnectivity metric based on the bivariate e relationship between each metric and P retention. Prior
studies have used a similar binary splitting w
 approach to examine the effect of various exogenous
factors on lake P retention (Cheng et al. 2010; Shimoda O and Arhonditsis 2015). We determined
n
splitting criteria for each metric from the results of a random forest procedure incorporating
ly
conditional inference trees (ctree, Hothorn and Zeileis 2015). This procedure creates binary splits
of the independent variables (i.e. each of the connectivity metrics), which are recursively
repeated to find the split that maximizes association with the dependent variable (P retention).
The advantage of the ctree technique over the more typical classification-regression tree (CART)
technique is that tree growth stopping rules are pre-specified (Hothorn et al., 2006). As a result,
some of subjectivity associated with post-hoc tree pruning is avoided.

@@

We examined the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention to determine the potential for strong relationships among
any of the variables which were not accounted for by our model structure (Eq 2, Figure 4). We
found evidence for some relationships among these variables, but none that suggest either
redundancy among connectivity metrics or that otherwise limit our ability to infer relationships
with P retention (Figure 4). For example, the Pearson correlation (r) between water residence
Ftime and stream density, o
 which is implicitly accounted for in our model structure, was 0.29 (p <
r0.05). In contrast, the correlation between water residence time and upstream lake area, which is
R
unaccounted for in our model structure, e was only 0.10 (p > 0.05). Note that interactions between
connectivity metrics and water residence v
 time are accounted for in our hierarchical model
istructure because Eq. 2 uses a global coefficient e x for water residence time that is estimated
separately from the hierarchical and connectivity-dependent w
 P decay coefficient k. This is
conceptually similar to fitting the k-connectivity-P retention O relationship to the residuals of the
n
water residence time to P retention relationship. Perhaps surprisingly, we did not observe strong
ly
correlations between P loading and water residence time (r = 0.13, p > 0.05) or between lake
depth and most connectivity metrics (r < 0.17). This suggests that our P retention results are not
confounded by variations in lake depth, by interactions between connectivity and P loading, or
by interactions between water residence time and P loading.
A secondary purpose of examining the bivariate relationships between connectivity,
water residence time, P loading and other factors related to P retention was to determine if our
connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among
connectivity metrics were low and of a similar magnitude as the correlations between
connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation
between any connectivity metric or lake characteristic was upstream lake area and lake
watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake
characteristics not accounted for in Eq 2. These include the correlation between P loading and
lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake
Fsubwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and
rupstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P
R
loading, watershed area, and upstream e lake area, we did not observe a strong correlation between
P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time e and P retention (r = 0.44, p > 0.05). Our
observation that the correlation between P w
 retention and P loading was not appreciably stronger
than correlations between P retention and our connectivity O metrics suggests that our estimates of
n
connectivity metric effects on P retention are not confounded by interactions between water
ly
residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found
that lakes with similar connectivity metric values, were spatially clustered (Figure 5). In
particular, we found that lakes were concentrated in either the southern or northern portions of
our study area depending on their connectivity metric value (Figure 5). This observation is
consistent with the findings of Fergus et al. (2017) that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our
study area.

@@

We found that freshwater connectivity metrics were associated with lake P retention (Figure 6,
7). Across all connectivity metrics except stream order ratio, we found that the P decay
coefficient k and thus P retention was associated with whether a lake had a high or low value of
each connectivity metric. These findings matched our expectations in several ways. Most
Fnotably, lakes with shorter o
 average link lengths had higher P retention relative to lakes with
rlonger average link lengths. In addition, lakes with less upstream lake area had higher P retention
R
than lakes with more upstream lake e area (Figure 6, 7).
We found that some connectivity v
 metrics were more strongly related to P retention than
iothers. For example, the model R2 for network e average link length was higher (R2 = 0.41) than
the global model (R2 = 0.34). For other connectivity w
 metrics, such as network stream order ratio,
goodness-of-fit (R2 = 0.36) was very similar to the global O model (R2 = 0.34). Model fit for other
n
connectivity metrics was in between these two extremes (0. 36 < R2 < 0.41). Overall, we found
ly
that all hierarchical models had at least a marginally better fit to the water residence time versus
P retention relationship than a global model which does not account for connectivity (Figure 6).
Although we found a discernable effect of connectivity metrics on lake P retention, the
somewhat modest improvements in model fit may be due to the fact that water residence time
remains a dominant effect even after accounting for freshwater connectivity.

@@

Differences in P retention among lakes with different connectivity metric values was reflected in
differences among connectivity class-specific values of the P decay parameter k (Figure 7). The
connectivity metric that had the most effect on k was average link length (Table 2). For instance,
hierarchical models fit with the average link length metric had a median effect size of 0.23 and
0.05 for k and P retention respectively (Table 2), which means that for this metric, lakes with
shorter average link lengths retained 4.7 to 4.9 % more P than lakes with longer average link
lengths. The influence of lake-stream-based connectivity metrics on lake P retention was similar
to stream-based connectivity metrics (Figure 7). This suggests that both lake-stream-based
Fconnectivity between lakes o
 and streams as well as stream-based connectivity of lakes and their
rterrestrial watersheds influence P retention (Figure 2).
R
We found that connectivity metrics e measured at the lake watershed extent were more
strongly associated with P retention v
 than metrics measured at the lake subwatershed extent
i(Table 2, Figure 7). Specifically, the metrics e that had the strongest association with P retention
such as average link length ($\delta$k = 0.23), closest w
 lake distance ($\delta$k = 0.22), and stream density ($\delta$k
= 0.20) also had a stronger association at the lake watershed O extent rather than at the lake
n
subwatershed extent (Table 2, Figure 7). An exception to this pattern was observed with the
ly
baseflow connectivity metric where although greater differences at the lake subwatershed extent
were more strongly associated with P retention, the sign of the effect was variable depending on
measurement extent (i.e. the value of the P decay parameter was positively related to
connectivity metric values at the subwatershed extent but was negatively related at the watershed
extent). For several connectivity metrics, we judged that differences in P decay among lakes with
either low or high connectivity metric values were not significant because they translated to
differences in P retention that were less than measurement precision (Table 2).

## Discussion
Although prior studies have found that lake P retention is related to water residence time (Brett
and Benjamin, Cheng, et al. 2010, Vollenweider 1975), there is substantial variation around this
relationship, particularly at intermediate water residence times. We found that some of this
remaining variation could be explained using a hierarchical modelling framework that accounts
for differences in freshwater connectivity among lakes. Although we found that the magnitude of
this effect depends on the specific connectivity metric, our results are consistent with the findings of previous studies o
 showing that connectivity metrics are associated with differences in
rlake carbon input fluxes and differences in lake nitrogen output fluxes (Cardille et al. 2007;
R
Schmadel et al. 2018). We also found e important differences in the association between
connectivity metrics and P retention v
 at different measurement extents. Specifically, we found
ithat P retention was more strongly associated e with connectivity measured at the broader lake
watershed extent rather than connectivity measured w
 at the finer lake subwatershed extent.

@@

We found that differences in P retention among lake connectivity classes was influenced by
specific connectivity metrics including average link length, closest lake distance, and stream
density (Table 2, Figure 7). Indeed, these metrics were more strongly associated with P retention
than covariates typically used in statistical P retention models (e.g. lake depth, Figure 1). Note
that we were able to examine the influence of specific metrics as separate effects apart from
water residence time because our model structure treats them as hierarchical coefficients on the P
decay parameter k. As a result, although water residence time remains the dominant effect on
lake P retention, we were able to estimate the specific influence of each metric in a way that is
not possible using integrated water residence time and connectivity metrics such as watershed
transport capacity (Fraterrigo and Downing 2008).
Several of the connectivity metrics that were most strongly associated with P retention
were weakly correlated with watershed size (Figure 4). The weak nature of these correlations are
consistent with the mixed results of prior studies linking watershed size to lake processes. For
example, Zimmer and McGlynn (2018) found that carbon export was related to watershed size,
but Smith et al. (2003) found that the nitrogen flux was not strongly related to watershed size.
FTaken together, our results o
 and the results of prior studies suggest that watershed size and lake
rdepth alone may not always reflect functional differences in potential material transport. One
R
consequence of a correlation between e connectivity metrics and watershed size is that metrics
derived from watershed size, such as v
 catchment to lake area ratio, are also likely to be associated
iwith connectivity. Catchment to lake area e ratio in particular has been previously used as an
approximation or proxy of water residence w
 time (Rasmussen et al. 1989; Sobek et al. 2007).
Although our results differ from those of Soranno et al. (2015) who found that the presence of an upstream lake connection was not strongly associated with catchment to lake area ratio, our results suggest that catchment to lake area ratio likely incorporates some connectivity information and caution is needed before using it as a proxy for water residence time.
Another lake characteristic that was strongly related to watershed size was P loading. In
particular, positive correlations between P loading and watershed size are consistent with the
idea that lakes in larger watersheds receive greater P loading (Prairie and Kalff 1986). A related
observation is that the correlation between P loading and upstream lake area was positive. This
can be explained by the fact that larger watersheds often have greater numbers of lakes nested within them (Zhang et al. 2012). Another notable result is that we observed a weaker correlation
between P loading and watershed area relative to the correlation between P loading and
subwatershed area. One explanation for this result is that land-use cover at the finer lake
subwatershed extent has more of an influence on P loading than land-use cover at the broader
lake watershed extent (Soranno et al. 2015).

@@

One of the challenges in quantifying the effect of freshwater connectivity on lake P retention is that it is not always known what spatial extent of the watershed is functionally connected to a lake. We found that connectivity at the broader lake watershed extent rather than connectivity at the finer lake subwatershed extent was e more strongly associated with differences in lake P
retention. In addition, we found that v
 individual characteristics such as lake depth (or more
idiscrete measures of connectivity such as e the presence of an upstream lake) were more strongly
associated with P retention than any of the w
 other connectivity metrics measured at the lake
subwatershed extent (Table 2). These findings contrast O with Soranno et al. (2015) who examined
n
the association between lake nutrient concentrations (as opposed to retention) and land-use
ly
measured at varying spatial extents and found that measurements at the finer lake subwatershed
extent rather than measurements at the broader lake watershed extent were more strongly
associated with lake nutrient concentrations. One explanation for the difference between our
results and those of Soranno et al. (2015) is that connectivity metrics may reflect long-range
watershed processes to a greater degree than land-use cover. An alternative explanation is that
controls of lake P retention may differ compared to controls on lake P concentration.

>>>>

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  texlive: true
  sysdeps: 
    - apt-utils
    - grass
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.4\textwidth,keepaspectratio]{../figures/02_conny_metric_key_crop_small} 
  \end{center}
  \caption{Connectivity metric definitions along with simplified examples of high and low value lakes that might arise from a binary classification. Both lake-stream-based and stream-based metrics are associated with restrictions on  in-stream transport whereas stream-based metrics are associated with differences in transport of P from terrestrial runoff to streams. We use the term "first order stream" to describe a headwater stream.}\label{fig2}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/03_iws_nws} 
  \end{center}
  \caption{Diagram showing the lake subwatershed and lake watershed of three lakes. Here the lake subwatershed of lake 3 encompasses the lake subwatershed of lake 2 because it is smaller than 10 ha but it does not encompass the lake subwatershed of lake 1 because it has an area of at least 10 ha. In contrast to the lake subwatershed boundaries, the lake watershed boundaries extend to the headwaters of the lake chain.}\label{fig3}
\end{figure}

@@

In addition to data from the NES and  LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related e to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing  aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo 2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we
expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal
to 10 ha (0.1 km2). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input (USGS e 2018). Average link length was calculated as the total stream length in a given watershed divided v
 by the number of stream reaches after dissolving (removing) any network points that do not e occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes)
expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams (Barbera and Rosso 1989). Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively (Stachelek 2018a, Stachelek 2018b). The algorithms in the `streamnet` package use the `sf` R package (Pebesma 2018) as well as the `v.net` and `v.stream.order` modules (Jasiewicz and Metz 2011) included in GRASS GIS (GRASS Development Team 2017). All processed connectivity data and code are available at [Zenodo DOI].

@@

where greater differences in k between R the two groups indicate greater support for a connectivity effect on P retention. Prior to model fitting, we examined the bivariate relationships between each connectivity metric, water residence i time, P loading, and other factors related to P retention
eusing Pearson’s correlation coefficients. The purpose of this exercise was twofold, to determine the potential for collinearity among any of the variables in Eq 2 and to identify any relationships between P retention, water residence time, and other watershed and lake characteristics that were not included in our model. As only one connectivity metric was used to define g for each model we did not use the results of this exercise to exclude variables from further investigation. We quantified the relative support for an effect of each connectivity metric on P retention in more detail by calculating the difference in the median value of the P decay parameter k between groups (i.e. $\Delta$k). We used these median k values along with median estimates of x to determine how differences in k translated to differences in P retention (Eq 2). We judged significance by whether or not differences in group-wise P retention were greater than the measurement precision of P retention (> 0.01).

We fit all models in a Bayesian framework using the non-linear extension to the brms package to interface with the Stan statistical program (Burkner 2017, Stan Development Team 2017). In both models, we set a semi-informative prior on k and x of N(1.3, 0.1) and N(0.45, 0.1) respectively. These priors were based on the confidence intervals presented in Brett and Benjamin (2007) and qualitatively matched those used by Cheng et al. (2010). We used the default settings of brms and rstan to generate posterior estimates using four chains of 4,000 iterations each with no thinning and initial parameter values drawn from a uniform distribution bounded between -2 and 2. We also used the brms package for model evaluation by computing a Bayesian $R^{2}$ following the method of Gelman et al. (2017).

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each connectivity metric based on the bivariate e relationship between each metric and P retention. Prior studies have used a similar binary splitting approach to examine the effect of various exogenous factors on lake P retention (Cheng et al. 2010; Shimoda O and Arhonditsis 2015). We determined splitting criteria for each metric from the results of a random forest procedure incorporating conditional inference trees (ctree, Hothorn and Zeileis 2015). This procedure creates binary splits of the independent variables (i.e. each of the connectivity metrics), which are recursively repeated to find the split that maximizes association with the dependent variable (P retention). The advantage of the ctree technique over the more typical classification-regression tree (CART) technique is that tree growth stopping rules are pre-specified (Hothorn et al., 2006). As a result, some of subjectivity associated with post-hoc tree pruning is avoided.

@@

We examined the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention to determine the potential for strong relationships among any of the variables which were not accounted for by our model structure (Eq 2, Figure 4). We found evidence for some relationships among these variables, but none that suggest either redundancy among connectivity metrics or that otherwise limit our ability to infer relationships with P retention (Figure 4). For example, the Pearson correlation (r) between water residence time and stream density, which is implicitly accounted for in our model structure, was 0.29 (p < 0.05). In contrast, the correlation between water residence time and upstream lake area, which is unaccounted for in our model structure, e was only 0.10 (p > 0.05). Note that interactions between connectivity metrics and water residence time are accounted for in our hierarchical model structure because Eq. 2 uses a global coefficient e x for water residence time that is estimated separately from the hierarchical and connectivity-dependent P decay coefficient k. This is conceptually similar to fitting the k-connectivity-P retention O relationship to the residuals of the water residence time to P retention relationship. Perhaps surprisingly, we did not observe strong correlations between P loading and water residence time (r = 0.13, p > 0.05) or between lake depth and most connectivity metrics (r < 0.17). This suggests that our P retention results are not confounded by variations in lake depth, by interactions between connectivity and P loading, or by interactions between water residence time and P loading.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/07_cor_mat_hmap-1} 
  \end{center}
  \caption{Correlation among connectivity metrics and selected lake characteristics. Cell shading and color reflects the correlation coefficient value. Only coefficients accompanied by a significant p value < 0.05 are shown as text. Here, WS is the lake watershed extent whereas SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure 2.}\label{fig4}
\end{figure}

A secondary purpose of examining the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention was to determine if our connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among connectivity metrics were low and of a similar magnitude as the correlations between connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation between any connectivity metric or lake characteristic was upstream lake area and lake watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake characteristics not accounted for in Eq 2. These include the correlation between P loading and lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake subwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and upstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P loading, watershed area, and upstream e lake area, we did not observe a strong correlation between P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time e and P retention (r = 0.44, p > 0.05). Our observation that the correlation between P retention and P loading was not appreciably stronger
than correlations between P retention and our connectivity O metrics suggests that our estimates of connectivity metric effects on P retention are not confounded by interactions between water residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found that lakes with similar connectivity metric values, were spatially clustered (Figure 5). In particular, we found that lakes were concentrated in either the southern or northern portions of our study area depending on their connectivity metric value (Figure 5). This observation is consistent with the findings of Fergus et al. (2017) that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our study area.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/08_maps-1} 
  \end{center}
  \caption{Locations of lakes of with different connectivity metric values at the (A) lake subwatershed and (B) lake watershed extent. Low valued lakes are represented by lighter (green) symbols, and high values lakes are represented by darker (purple) symbols. WS is the lake watershed extent. SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure 2.}\label{fig5}
\end{figure}

@@

We found that freshwater connectivity metrics were associated with lake P retention (Figure 6, 7). Across all connectivity metrics except stream order ratio, we found that the P decay coefficient k and thus P retention was associated with whether a lake had a high or low value of each connectivity metric. These findings matched our expectations in several ways. Most
Fnotably, lakes with shorter average link lengths had higher P retention relative to lakes with longer average link lengths. In addition, lakes with less upstream lake area had higher P retention than lakes with more upstream lake area (Figure 6, 7).

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/04_global_vollenweider-1} 
  \end{center}
  \caption{Residence time (yr) versus P retention for the NES dataset and the global model fit to the data ($R^{2}$ = 0.34, n = 129) where the solid line and shaded interval represents the median and central 95% interval estimates respectively (A). As above except that solid lines and shaded interval estimates represent hierarchical model fits to the data ($R^{2}$ = 0.41, n = 129) based on watershed average link length, which had the strongest association of any connectivity metric relative to P retention (B). Equations for these lines at median water residence time for low and high link length lakes are: Rp = 1 - (1 / (1 + 1.32 $\protect\tau^{0.4}$)) and Rp = 1 - (1 / (1 + 1.08 $\protect\tau^{0.4}$)) respectively.}\label{fig6}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/06_k-3} 
  \end{center}
  \caption{Distribution of the k parameter from Eq. 1 in lakes of differing connectivity, depth, or baseflow at the (A) lake subwatershed and (B) lake watershed extents. Lighter (green) lines indicate lakes with lower connectivity metrics values while darker (purple) lines indicate lakes with higher connectivity metrics values. For lake connection, lighter colored lines indicate lakes without upstream lakes. Connectivity metrics are defined in Figure 2. Labels associated with models where differences in k translated to significant differences in P retention are bolded and starred.}\label{fig7}
\end{figure}

We found that some connectivity metrics were more strongly related to P retention than others. For example, the model $R^{2}$ for network e average link length was higher ($R^{2}$ = 0.41) than the global model ($R^{2}$ = 0.34). For other connectivity metrics, such as network stream order ratio,
goodness-of-fit ($R^{2}$ = 0.36) was very similar to the global O model ($R^{2}$ = 0.34). Model fit for other connectivity metrics was in between these two extremes (0. 36 < $R^{2}$ < 0.41). Overall, we found that all hierarchical models had at least a marginally better fit to the water residence time versus P retention relationship than a global model which does not account for connectivity (Figure 6).
Although we found a discernable effect of connectivity metrics on lake P retention, the somewhat modest improvements in model fit may be due to the fact that water residence time remains a dominant effect even after accounting for freshwater connectivity.

@@

Differences in P retention among lakes with different connectivity metric values was reflected in differences among connectivity class-specific values of the P decay parameter k (Figure 7). The connectivity metric that had the most effect on k was average link length (Table 2). For instance, hierarchical models fit with the average link length metric had a median effect size of 0.23 and 0.05 for k and P retention respectively (Table 2), which means that for this metric, lakes with shorter average link lengths retained 4.7 to 4.9 % more P than lakes with longer average link lengths. The influence of lake-stream-based connectivity metrics on lake P retention was similar
to stream-based connectivity metrics (Figure 7). This suggests that both lake-stream-based connectivity between lakes and streams as well as stream-based connectivity of lakes and their terrestrial watersheds influence P retention (Figure 2).

We found that connectivity metrics e measured at the lake watershed extent were more strongly associated with P retention than metrics measured at the lake subwatershed extent (Table 2, Figure 7). Specifically, the metrics e that had the strongest association with P retention such as average link length ($\Delta$k = 0.23), closest lake distance ($\Delta$k = 0.22), and stream density ($\Delta$k = 0.20) also had a stronger association at the lake watershed O extent rather than at the lake subwatershed extent (Table 2, Figure 7). An exception to this pattern was observed with the baseflow connectivity metric where although greater differences at the lake subwatershed extent
were more strongly associated with P retention, the sign of the effect was variable depending on measurement extent (i.e. the value of the P decay parameter was positively related to connectivity metric values at the subwatershed extent but was negatively related at the watershed extent). For several connectivity metrics, we judged that differences in P decay among lakes with either low or high connectivity metric values were not significant because they translated to differences in P retention that were less than measurement precision (Table 2).

## Discussion

Although prior studies have found that lake P retention is related to water residence time (Brett and Benjamin, Cheng, et al. 2010, Vollenweider 1975), there is substantial variation around this relationship, particularly at intermediate water residence times. We found that some of this remaining variation could be explained using a hierarchical modelling framework that accounts for differences in freshwater connectivity among lakes. Although we found that the magnitude of this effect depends on the specific connectivity metric, our results are consistent with the findings of previous studies  showing that connectivity metrics are associated with differences in lake carbon input fluxes and differences in lake nitrogen output fluxes (Cardille et al. 2007; Schmadel et al. 2018). We also found e important differences in the association between connectivity metrics and P retention at different measurement extents. Specifically, we found that P retention was more strongly associated e with connectivity measured at the broader lake watershed extent rather than connectivity measured at the finer lake subwatershed extent.

@@

We found that differences in P retention among lake connectivity classes was influenced by specific connectivity metrics including average link length, closest lake distance, and stream density (Table 2, Figure 7). Indeed, these metrics were more strongly associated with P retention than covariates typically used in statistical P retention models (e.g. lake depth, Figure 1). Note that we were able to examine the influence of specific metrics as separate effects apart from water residence time because our model structure treats them as hierarchical coefficients on the P decay parameter k. As a result, although water residence time remains the dominant effect on lake P retention, we were able to estimate the specific influence of each metric in a way that is not possible using integrated water residence time and connectivity metrics such as watershed transport capacity (Fraterrigo and Downing 2008).
Several of the connectivity metrics that were most strongly associated with P retention were weakly correlated with watershed size (Figure 4). The weak nature of these correlations are consistent with the mixed results of prior studies linking watershed size to lake processes. For example, Zimmer and McGlynn (2018) found that carbon export was related to watershed size,
but Smith et al. (2003) found that the nitrogen flux was not strongly related to watershed size.
Taken together, our results and the results of prior studies suggest that watershed size and lake depth alone may not always reflect functional differences in potential material transport. One consequence of a correlation between e connectivity metrics and watershed size is that metrics derived from watershed size, such as catchment to lake area ratio, are also likely to be associated with connectivity. Catchment to lake area e ratio in particular has been previously used as an approximation or proxy of water residence w
 time (Rasmussen et al. 1989; Sobek et al. 2007).

Although our results differ from those of Soranno et al. (2015) who found that the presence of an upstream lake connection was not strongly associated with catchment to lake area ratio, our results suggest that catchment to lake area ratio likely incorporates some connectivity information and caution is needed before using it as a proxy for water residence time.
Another lake characteristic that was strongly related to watershed size was P loading. In particular, positive correlations between P loading and watershed size are consistent with the idea that lakes in larger watersheds receive greater P loading (Prairie and Kalff 1986). A related observation is that the correlation between P loading and upstream lake area was positive. This can be explained by the fact that larger watersheds often have greater numbers of lakes nested within them (Zhang et al. 2012). Another notable result is that we observed a weaker correlation between P loading and watershed area relative to the correlation between P loading and subwatershed area. One explanation for this result is that land-use cover at the finer lake subwatershed extent has more of an influence on P loading than land-use cover at the broader lake watershed extent (Soranno et al. 2015).

@@

One of the challenges in quantifying the effect of freshwater connectivity on lake P retention is that it is not always known what spatial extent of the watershed is functionally connected to a lake. We found that connectivity at the broader lake watershed extent rather than connectivity at the finer lake subwatershed extent was more strongly associated with differences in lake P retention. In addition, we found that individual characteristics such as lake depth (or more discrete measures of connectivity such as e the presence of an upstream lake) were more strongly associated with P retention than any of the other connectivity metrics measured at the lake subwatershed extent (Table 2). These findings contrast O with Soranno et al. (2015) who examined the association between lake nutrient concentrations (as opposed to retention) and land-use measured at varying spatial extents and found that measurements at the finer lake subwatershed extent rather than measurements at the broader lake watershed extent were more strongly associated with lake nutrient concentrations. One explanation for the difference between our results and those of Soranno et al. (2015) is that connectivity metrics may reflect long-range watershed processes to a greater degree than land-use cover. An alternative explanation is that controls of lake P retention may differ compared to controls on lake P concentration.
